---
title: "cirrhosisPrediction"
author: "Kanishk Kapoor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this project is to analyze patient data and develop a predictive model to determine which factors influence survival outcomes in cirrhosis patients.
```{r}
library(dplyr)
library(survival)
library(survminer)
library(ggplot2)
```

```{r}
c_data <- read.csv("cirrhosis.csv", stringsAsFactors = FALSE)
```

```{r}
View(c_data)
```

```{r}
str(c_data)
```

```{r}
colSums(is.na(c_data))
```

Handling Missing Categorical Data

```{r}
c_data$Drug[is.na(c_data$Drug)] <- "Unknown"
c_data$Ascites[is.na(c_data$Ascites)] <- "Unknown"
c_data$Hepatomegaly[is.na(c_data$Hepatomegaly)] <- "Unknown"
c_data$Spiders[is.na(c_data$Spiders)] <- "Unknown"
```
```{r}
colSums(is.na(c_data))
```

Handling Missing Numerical Data

```{r}
median_impute <- function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)
  return(x)
}

c_data$Cholesterol <- median_impute(c_data$Cholesterol)
c_data$Copper <- median_impute(c_data$Copper)
c_data$Alk_Phos <- median_impute(c_data$Alk_Phos)
c_data$SGOT <- median_impute(c_data$SGOT)
c_data$Tryglicerides <- median_impute(c_data$Tryglicerides)
c_data$Platelets <- median_impute(c_data$Platelets)
c_data$Prothrombin <- median_impute(c_data$Prothrombin)


```
```{r}
c_data$Stage[is.na(c_data$Stage)] <- 0
c_data$Stage <- as.factor(c_data$Stage)
```

```{r}
str(c_data)
colSums(is.na(c_data))
```

0 in stage to imply unknown value.
```{r}
#from days age to years
c_data$Age <- c_data$Age / 365
```
```{r}
str(c_data)
```

```{r}
categorical_cols <- c("Status", "Drug", "Sex", "Ascites", "Hepatomegaly", "Spiders", "Edema", "Stage")
c_data[categorical_cols] <- lapply(c_data[categorical_cols], as.factor)
```

### Exploratory Data Analysis
```{r}
summary(c_data)
```

```{r}
hist(c_data$N_Days, breaks = 30, main="Distribution of Survival Time", col = "blue")
```

#### Surival Analysis using Kaplan Meier Curve

```{r}
table(c_data$Drug)
str(c_data$Drug)
```
```{r}
c_data_s <- c_data %>% filter(c_data$Drug != "Unknown")
table(c_data_s$Drug)
```

```{r}
surv_obj <- Surv(c_data_s$N_Days, c_data_s$Status =="D")

km_fit <- survfit(surv_obj ~ Drug, data = c_data_s)

```
```{r}
ggsurvplot(km_fit, data = c_data_s, pval = TRUE, risk.table = TRUE, main = "Survival Probability by Drug Type", legend.title= "Drug")
```
Key Observations
	1.	Survival Curves are Close Together
	•	The two lines are overlapping for most of the time period.
	•	This suggests no significant survival benefit from D-penicillamine compared to Placebo.
	2.	p-value = 0.75
	•	A p-value > 0.05 means no statistically significant difference between the two groups.
	•	This means D-penicillamine does not significantly improve survival over Placebo.
	3.	Long-term Survival Patterns
	•	Both curves decline gradually, indicating a progressive disease outcome.
	•	The survival probability at 4000+ days is low for both groups.

Checking for Other Factors which may impact survival

```{r}

km_fit_stage <- survfit(surv_obj ~ Stage, data=c_data_s)
ggsurvplot(km_fit_stage,
           data= c_data_s,
           pval= TRUE,
           risk.table = TRUE,
           main="Survival Probability by Cirrhosis Stage",
           legend.title="Stage")
```
stage cirrhosis have the best survival rates.
	•	Stage 2 (Green Line): Slightly lower than Stage 1 but still better than advanced stages.
	•	Stage 3 (Blue Line): Noticeably lower survival compared to Stages 1 and 2.
	•	Stage 4 (Purple Line): Worst survival rate—a sharp decline in survival probability, indicating that patients with advanced cirrhosis (Stage 4) have the highest mortality risk.

Key Takeaways:

✔️ The survival probability decreases as the cirrhosis stage increases.
✔️ p < 0.0001 suggests that the difference in survival across stages is statistically significant.
✔️ Medical Implication: Early detection and treatment of cirrhosis can significantly improve patient survival.
```{r}
km_fit_sex <- survfit(surv_obj ~ Sex, data=c_data_s)
ggsurvplot(km_fit_sex,
           data= c_data_s,
           pval= TRUE,
           risk.table = TRUE,
           main="Survival Probability by Gender",
           legend.title="Sex")
```
Key Observations:

1️⃣ The red line (females) stays consistently higher than the blue line (males), indicating longer survival.
2️⃣ The gap between curves widens over time, suggesting that males experience faster mortality rates.
3️⃣ p = 0.039, which is statistically significant (below 0.05), meaning the difference in survival between males and females is unlikely to be due to chance.
4️⃣ Number at risk decreases faster for males, reinforcing that they have a higher risk of mortality.

Possible Medical Explanation:
	•	Women may have better liver regeneration or respond differently to cirrhosis-related treatments.
	•	Hormonal differences could play a role in disease progression.
	•	Males may have more aggressive disease progression due to lifestyle factors (e.g., alcohol consumption).
```{r}
km_fit_Ascites<- survfit(surv_obj ~ Ascites, data=c_data_s)
ggsurvplot(km_fit_Ascites,
           data= c_data_s,
           pval= TRUE,
           risk.table = TRUE,
           main="Survival Probability by Ascites",
           legend.title="Ascites")
```
Key Observations:

1️⃣ The blue line (Ascites = Yes) drops sharply, indicating much lower survival rates for patients with Ascites.
2️⃣ The red line (Ascites = No) remains consistently higher, meaning patients without Ascites survive significantly longer.
3️⃣ p < 0.0001, which is highly statistically significant, confirming that the difference in survival is not due to random chance.
4️⃣ The number at risk decreases much faster for patients with Ascites (24 → 0 quickly), while those without it still have surviving patients even at 4000 days.

Interpretation:
	•	Ascites is a serious complication of cirrhosis, associated with high mortality.
	•	Patients with Ascites might have worse liver function or higher risk of liver failure.
	•	These findings suggest Ascites should be a key predictor in a Cox regression model.
	
```{r}
km_fit_Hepatomegaly<- survfit(surv_obj ~ Hepatomegaly, data=c_data_s)
ggsurvplot(km_fit_Hepatomegaly,
           data= c_data_s,
           pval= TRUE,
           risk.table = TRUE,
           main="Survival Probability by Hepatomegaly",
           legend.title="Hepatomegaly")
```
Key Observations:

1️⃣ The blue line (Hepatomegaly = Yes) drops faster, meaning patients with Hepatomegaly have poorer survival rates.
2️⃣ The red line (Hepatomegaly = No) remains consistently higher, indicating better survival for patients without Hepatomegaly.
3️⃣ p < 0.0001, meaning the difference is highly statistically significant and unlikely due to random variation.
4️⃣ The number at risk table shows that survival decreases more rapidly in the Hepatomegaly = Yes group.

Interpretation:
	•	Hepatomegaly is a sign of liver disease progression, often indicating advanced cirrhosis, congestion, or liver dysfunction.
	•	The survival curve for patients with Hepatomegaly declines sharply, suggesting it could be a key predictor in a Cox model.
	
```{r}
km_fit_Spiders<- survfit(surv_obj ~ Spiders, data=c_data_s)
ggsurvplot(km_fit_Spiders,
           data= c_data_s,
           pval= TRUE,
           risk.table = TRUE,
           main="Survival Probability by Spiders",
           legend.title="Spiders")
```
	
```{r}
km_fit_Edema<- survfit(surv_obj ~ Edema, data=c_data_s)
ggsurvplot(km_fit_Edema,
           data= c_data_s,
           pval= TRUE,
           risk.table = TRUE,
           main="Survival Probability by Edema",
           legend.title="Edema")
```
#### Cox Proportional Hazards Model
```{r}
 c_data_s$Status_B <- ifelse(c_data_s$Status == "D", 1, 0)
table(c_data_s$Status_B)  # Should now contain only 0 and 1
```


```{r}
#Univariate Cox Model

cox_stage <- coxph(Surv(N_Days, Status_B) ~ Stage, data= c_data_s)
cox_ascites <- coxph(Surv(N_Days, Status_B) ~ Ascites, data = c_data_s)
cox_hepatomegaly <- coxph(Surv(N_Days, Status_B) ~ Hepatomegaly, data = c_data_s)
cox_spiders <- coxph(Surv(N_Days, Status_B) ~ Spiders, data = c_data_s)
cox_edema <- coxph(Surv(N_Days, Status_B) ~ Edema, data = c_data_s)

cox_bilirubin <- coxph(Surv(N_Days, Status_B) ~ Bilirubin, data=c_data_s)
cox_albumin <- coxph(Surv(N_Days, Status_B) ~ Albumin, data = c_data_s)
cox_platelets <- coxph(Surv(N_Days, Status_B) ~ Platelets, data = c_data_s)

summary(cox_stage)
summary(cox_ascites)
summary(cox_hepatomegaly)
summary(cox_spiders)
summary(cox_edema)
summary(cox_bilirubin)
summary(cox_albumin)
summary(cox_platelets)


```

```{r}
multivariate_model <- coxph(Surv(N_Days, Status_B) ~ Stage + Ascites + Hepatomegaly + Spiders + Edema + Bilirubin + Albumin + Platelets, data = c_data_s)
summary(multivariate_model)
```

Significant Predictors (p < 0.05)
	1.	Stage (Stage1 & Stage2)
	•	Stage1: HR = 0.13 (p = 0.048) → Patients in Stage1 have a significantly lower risk of death compared to Stage4.
	•	Stage2: HR = 0.49 (p = 0.026) → Patients in Stage2 also have a reduced risk compared to Stage4.
	2.	Edema
	•	EdemaY: HR = 2.24 (p = 0.016) → Presence of edema increases the hazard of death by 2.24 times.
	3.	Bilirubin
	•	HR = 1.13 (p < 0.001) → A 1-unit increase in bilirubin is associated with a 12.7% increase in the hazard of death.
	4.	Albumin
	•	HR = 0.44 (p = 0.0026) → A 1-unit increase in albumin decreases the risk of death by 56%.

Non-Significant Predictors (p > 0.05)
	•	Ascites (p = 0.070): Potentially significant, but not quite below the 0.05 threshold.
	•	Hepatomegaly, Spiders, Edema (S), Platelets: Not significant.

⸻

Model Performance
	•	Concordance Index = 0.823 → Strong predictive power (closer to 1 is better).
	•	Likelihood Ratio Test, Wald Test, Log-rank Test: All p < 2e-16 → Model is highly significant.
	
	 Disease Stage is a Strong Predictor
	•	Patients in Stage 1 have the lowest risk (HR = 0.1299), meaning 87% lower risk compared to Stage 4.
	•	Higher stages (Stage 2 & 3) also show increased risk but not as extreme as Stage 4.

#### Conclusion

 Ascites & Edema Significantly Impact Survival
	•	Ascites (HR = 1.76) → 76% higher risk of death.
	•	Edema (HR = 2.24) → More than double the risk of death.

 Bilirubin is a Major Risk Factor
	•	Each unit increase in bilirubin raises the death risk by 13% (HR = 1.13, p < 0.0001).
	•	Higher bilirubin levels indicate poorer survival.

 Albumin is Protective
	•	Higher albumin levels reduce risk (HR = 0.44, p = 0.0026).
	•	Suggests better liver function is linked to longer survival.

 Platelets & Hepatomegaly have weaker effects
	•	Platelets show a weak association with survival (HR = 0.9987, p = 0.203).
	•	Hepatomegaly is not statistically significant in predicting survival.